---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path='figure/',
                      warning=FALSE, message=FALSE)
```

## Step 1. Loading and preprocessing the data

The dataset is stored in activity.csv (unpacked from activity.zip) and there are 17.568 observations. 
The following code unpacks the file in "data folder"" and loads in "data" variable.

The variables included in research dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken


```{r}
  if (!file.exists("./data")) {
    dir.create("./data");
  }
  if (!file.exists("./data/activity.csv")) {
    unzip("activity.zip", exdir = "data");
  }  
  data <- read.csv("./data/activity.csv")

```

Number of rows in the given dataset:
```{r echo=FALSE}

nrow(data)

```


## Step 2. What is mean total number of steps taken per day?

The following code calculates total number if steps per day:
```{r}
  maindata <- aggregate(. ~ date, data = data, sum)
  maindata[, c("date", "steps")]

```

Histogram of the disribution of number of steps taken per day:
```{r steps_per_day_hist, warning=FALSE}
  library(ggplot2)
  ggplot(maindata, aes(x=steps)) +
    geom_histogram(colour="black", fill="blue")
```

Mean and Median of steps per day:
```{r}
  mean(maindata$steps)
  median(maindata$steps)
```

## Step 3. What is the average daily activity pattern?

Averages of steps for each time interval across days:
```{r}
  intraday <- aggregate(. ~ interval, data= data, mean)[,c("interval", "steps")]
```

Plot shows distribution of number of steps over the day:
```{r steps_per_5minutes_dist, warning=FALSE}
  ggplot(intraday, aes(x=interval, y=steps)) + geom_line(colour="green")
```

Maximum number of steps was made in the following 5-minute interval:
```{r}
  intraday[max(intraday$steps) == intraday$steps, "interval"]

```

## Imputing missing values

Amount of rows with NA:
```{r}
sum(is.na(data$steps))
```

Filling missing values with average of steps for same 5-minute intervals on other day:
```{r}
  fulldata <- merge(x=data, y=intraday, by="interval", all.x = TRUE)
  fulldata[is.na(fulldata$steps.x),"steps.x"] <- fulldata[is.na(fulldata$steps.x),"steps.y"]
```

Aggregation data by day:
```{r}
  maindataFull <- aggregate(. ~ date, data = fulldata, sum)[, c("date", "steps.x")]

```

Histogram of the disribution of amount of steps:
```{r steps_per_day_hist_full, warning=FALSE}
  ggplot(maindataFull, aes(x=steps.x)) +
    geom_histogram(colour="black", fill="blue")
```

Mean and Median of steps per day:
```{r}
  mean(maindataFull$steps.x)
  median(maindataFull$steps.x)
```

The shape of histogram has not changed much, some bars have higher values. However, median and mean values are equals now what could be explained by usage of average for 5-minutes interval to replace empty values.

## Are there differences in activity patterns between weekdays and weekends?

Creating wday xolumn with weekday/weekend factor:
```{r}
fulldata$wday <- factor(ifelse(weekdays(as.Date(fulldata$date)) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

Computing average data by 5-minutes intervals:
```{r}
intradayWithDays <- aggregate(steps.x ~ wday + interval, data = fulldata, mean)
```

Results on plot:
```{r steps_per_day_wday}
ggplot(intradayWithDays, aes(x=interval, y=steps.x)) + 
    geom_line(colour="blue") +
    facet_grid(wday ~ .) +
    ylab("Steps by 5-minutes interval") +
    ggtitle("Step distribution by weekday and weekend")
```
